Predicting the PCG PSeudo-Random Number Generator In Practice

### 2020-05-22 article/main.tex
algo c known v1.

\section{Reconstructing Truncated Geometric Sequences}
\todo{à fusionnier avec la section précédente!}
\subsection{The lattice}

@@

In this case, reconstructing the 128-bit internal state $S_i$ of the generator
obis sufficient to produce the pseudo-random flow with 100\% accuracy (the
generator can also be clocked backwards if necessary, so that the seed can be
easily reconstructed). We therefore focus on $S_0$, the first internal state
after seeding. A very simple strategy could be the following:
\begin{enumerate}
\item Guess the 64 upper bits of $S_0$.
\item Compute the missing 64 lower bits using~\eqref{eq:output}, with:
\[
   S_0[0:64] = S_0[64:128] \oplus (X_0  \lll S[122:128]).
\]
\item If $X_1$ is computed correctly, output $S_0$.
\end{enumerate}

@@

In figure~\ref{fig:Cknown}, we see that guessing $S_0[0:\ell]$ and a few 6-bit
rotations give access to $S_i[58:64+\ell]$ for the corresponding consecutive
states. Therefore, we are facing a truncated linear congruential generator on
$64+\ell$ bits, where we only know $6+\ell$ bits of output, for a few
consecutive outputs. This is sufficient to reconstruct entirely the successive
states of this truncated linear congruential generator. This reveals
$S_0[0:64+\ell]$, and using~\eqref{eq:output} the entire $S_0$ can be
reconstructed. The precise details follow.

@@

  % T_0
  \begin{scope}[xshift=4.6cm, yshift=-1.5cm]    
    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
    \draw[thick]  (0, 0) rectangle +(2.4, 1);
    \draw[]  (1.8, 0) rectangle +(0, 1);
    \path (2.4, 0) rectangle node {$00000000 \cdots 00000000$} (6.4, 1);

@@

  % T_1
  \begin{scope}[xshift=4.6cm, yshift=-4.5cm]    
    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
    \draw[thick]  (0, 0) rectangle +(2.4, 1);
    \draw[]  (1.8, 0) rectangle +(0, 1);
    \path (2.4, 0) rectangle node {$00000000 \cdots 00000000$} (6.4, 1);

@@

  % T_2
  \begin{scope}[xshift=4.6cm, yshift=-7.5cm]    
    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
    \draw[thick]  (0, 0) rectangle +(2.4, 1);
    \draw[]  (1.8, 0) rectangle +(0, 1);
    \path (2.4, 0) rectangle node {$00000000 \cdots 00000000$} (6.4, 1);

@@

\end{center}
\caption{A guess-and-determine algorithm to reconstruct the first internal state
  $S_0$. Magenta bits are guessed; cyan bits are obtained using the linear
  congruence relation~\eqref{eq:lcg}; yellow bits are obtained
  from the output and the guessed rotations using~\eqref{eq:output}.}
\label{fig:Cknown}
\end{figure}

Set $S_i' = S_i - (S_0[0:\ell] + c) \sum_{j=0}^i a^i$. It is easy to check that
$S'_{i+1} = a S'_i$, and therefore the successive $S'_i$'s follow a geometric
progression\footnote{note de bas de page : le fait de retirer $S_0[0:\ell]$
  n'est pas nécéssaire, mais nous permet de limiter la taille des instances
  utilisées à 64 bits, très utile pour la programmation en C)}.

Nous utiliserons le réseau engendré par les $S'[\ell : 64 + \ell]$, nommé $\mathcal{L}(A,2^64,n)$. On se réduit donc au cas facile étudié dans la partie précédente.
\begin{equation}
genMatrix =
\begin{pmatrix} 
1 & A & A ^2 & \dots & A ^{n- 1}\\
0 & M & 0 & \dots & 0\\
0 & 0 & M & \dots & 0\\
\vdots & & & \ddots & \vdots\\
0 & 0 & \dots & 0 & M
\end{pmatrix}
\end{equation}

On calcule préalablement les matrices $G$ et $G^{-1}$, avec $G$ la matrice génératrice réduite du réseau $\mathcal{L}(A,2^64,n)$ par l'algorithme LLL.
Pour tous les $r_i$ et les $S_0[0 : \ell]$ possibles:\\
\begin{enumerate}
\item On récupère le $T$ correspondant à nos suppositions. On calcule $T'$. ($T'_i = (S_i[64 - 6 : 64 + \ell] - (A^i*S_0[0 : \ell] + C * \sum_{j = 0}^i A^j)[64 - 6 : 64 + \ell])[64 - 6 : 64 + \ell] * 2^{64 - 6 - \ell}$)

\todo{schema T et T'}

Notons cependant que les composantes en $C$ et $S[0:l]$ sont calculées en dehors de la boucle et stockées dans un tableau pour des raisons de performance. Ceci est fait de même dans tous les algorithmes suivants.

\item We get the correspondant $S'[\ell : 64 + \ell]$ by using the babai rounding technique on $T'$ using $G$ and $G^{-1}$.

\item We test our solution by computing $S_0$ and testing if the following internals states are coherent with the PCG outputs. If the results are coherents, we return $S_0$.

\end{enumerate}

>>>>

\section{Reconstructing Truncated Geometric Sequences}
\label{sec:geometric}

\todo{à fusionnier avec la section précédente!}
\subsection{The lattice}

@@

In this case, reconstructing the 128-bit internal state $S_i$ of the generator
is sufficient to produce the pseudo-random flow with 100\% accuracy (the
generator can also be clocked backwards if necessary, so that the seed can be
easily reconstructed). We therefore focus on reconstructing $S_0$ (the seed)
from $X_0, X_1, X_2, \dots$. A very simple strategy could be the following:
\begin{enumerate}
\item Guess the 64 upper bits of $S_0$.
\item Compute the missing 64 lower bits using~\eqref{eq:output}, with:
\[
   S_0[0:64] = S_0[64:128] \oplus (X_0  \lll S[122:128]).
\]
\item If $X_1$ is computed correctly, output $S_0$.
\end{enumerate}

@@

In figure~\ref{fig:Cknown}, we see that guessing $S_0[0:\ell]$ and a few 6-bit
rotations give access to $S_i[58:64+\ell]$ for the corresponding
states. Therefore, looking at $S_i[\ell:64+\ell]$, we are facing a truncated
linear congruential generator on $64$ bits, where we have access to the most
$6+\ell$ bits of each state, for a few consecutive states. This is sufficient to
reconstruct entirely the successive states of this truncated linear congruential
generator. This reveals $S_0[\ell:64+\ell]$, and using~\eqref{eq:output} the
entire $S_0$ can be reconstructed. The precise details follow.

@@

  % T_0
  \begin{scope}[xshift=4.6cm, yshift=-1.5cm]    
    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
    \draw[thick]  (0, 0) rectangle +(2.4, 1);
    \draw[]  (1.8, 0) rectangle +(0, 1);
    \path (2.4, 0) rectangle node {$???????? \cdots ????????$} (6.4, 1);

@@

  % T_1
  \begin{scope}[xshift=4.6cm, yshift=-4.5cm]    
    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
    \draw[thick]  (0, 0) rectangle +(2.4, 1);
    \draw[]  (1.8, 0) rectangle +(0, 1);
    \path (2.4, 0) rectangle node {$???????? \cdots ????????$} (6.4, 1);

@@

  % T_2
  \begin{scope}[xshift=4.6cm, yshift=-7.5cm]    
    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
    \draw[thick]  (0, 0) rectangle +(2.4, 1);
    \draw[]  (1.8, 0) rectangle +(0, 1);
    \path (2.4, 0) rectangle node {$???????? \cdots ????????$} (6.4, 1);

@@

\end{center}
\caption{A guess-and-determine algorithm to reconstruct the first internal state
  $S_0$. Magenta bits are guessed; cyan bits are obtained using the linear
  congruence relation~\eqref{eq:lcg} modulo $2^\ell$; yellow bits are obtained
  from the output and the guessed rotations using~\eqref{eq:output}.}
\label{fig:Cknown}
\end{figure}

Set $S_i' = S_i - S_0[0:\ell] a^i - c \sum_{j=0}^{i-1} a^i$. The sequence
$(S'_i)$ describe the internal states we would observe if $c=0$ and
$S_0[0:\ell] = 0$. It is easy to check that $S'_{i} = a^i (S_0 - S_0[0:\ell])$,
and therefore the successive $S'_i$'s follow a geometric progression. % Indeed :
% \begin{align*}
% S_i &= a S_{i-1} + c\\
%     &= a^2 S_{i-2} + (a + 1) c \\
%     &\vdots  \\
%     &= a^i S_0  + c \sum_{j=0}^{i-1} a^i \\
%   S'_i  &= S_i - c \sum_{j=0}^{i-1} a^i - S_0[0:\ell] a^i \\
%     &= a^i (S_0 - S_0[0:\ell])
% \end{align*}
The $\ell$ low-order bits of $S_0 - S_0[0:\ell]$ are all zero by construction;
therefore set $T_i = \left(S'_i / 2^\ell\right) \bmod 2^{64}$ (the division by
$2^\ell$ is exact). It follows that the $(T_i)$ is also a geometric progression
with common ratio $a$, and we have access to the $\ell+6$ high order bits of
$T_i$ if we guess the $i$-th rotation. We have seen in
section~\ref{sec:geometric} that 19 known bits on 3 consecutive outputs are
sufficient for a deterministic reconstruction of the full sequence. Therefore,
we guess 19 low-order bits of the state and three consecutive rotations.

Therefore, the algorithm that reconstructs the internal state of the
\textsf{PCG64} generator with known increment proceeds as follows:

\begin{center}\noindent\framebox{
  \begin{minipage}{0.95\textwidth}
\aalgbegin Algorithm K (State reconstruction with known increment). Reconstruct $S_0$ from three consecutive outputs $X_j$ ($0 \leq j < 3$). In all the algorithm, we have $j= 0, 1, 2$.

\algstep K1. [Guess state] For each $0 \leq w < 2^{19}$, set $S_0[0 : \ell] = w$ and do steps K2--K10.

\algstep K2. [Propagate state] Set $S_j[0:\ell] \gets a^j w \bmod 2^{19}$.

\algstep K3. [Guess rotations] For each $0 \leq r_0, r_1, r_2 < 64$, do steps K4--K10.

\algstep K4. [Undo rotations] Set $S_j[122:128] \gets r_j$ and $Y_j \gets X_j \lll r_j$.

\algstep K5. [Propagate] Set $S_j[58:64] \gets r_j \oplus Y_j[58:64]$ and $S_j[64:83] \gets S_j[0:19] \oplus Y_j[0:19]$.

\algstep K6. [Build partial $T$] Set $T_j \gets \left(S_j[58:83] - (a^j w + c \sum_{k = 0}^j a^k)[58:83]\right) \cdot 2^{39}$

\algstep K7. [Reconstruct $T$] Set $(U_0, U_1, U_2) \gets \left\lfloor (T_0, T_1, T_2) \cdot \widetilde G^{-1} \right\rceil \cdot \widetilde G$.

\algstep K8. [Reconstruct $S_0$] Set $S_0[19:83] \gets U_0$ and $S_0[83:122] \gets U_0[0:39] \oplus Y_0[19:58]$.

\algstep K9. [Recompute $X_1$] Compute $\widehat S_1 \gets a S_0 + c$ and $\widehat X_1 =(\widehat S_1[0:64] \oplus \widehat S_1[64:128]) \ggg r_1$.

\algstep K10. [Check] If $\widehat{X}_1 = X_1$, the output $S_0$ as a candidate internal state.
\end{minipage}
}
\end{center}
